Predictions From Day 0 Community Structures

### 2016-08-09 code/HumanCdGF_microbiota_predict.Rmd
Create for loop RF by log10cfu day

# read in files
meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 'group')
shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header = T, row.names='Group')

# add log transformed CFU
meta_file$log_cfu <- log10(meta_file$cdiff_cfu + 1)

# remove cecum samples
meta_file <- meta_file[! meta_file$sample_type == 'cecum', ]

# subset shared
#Create df with relative abundances
stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
# use %in% instead of == because to ignore NA
# rename day 0 communities by mouse id
rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% rownames(stool_shared) ]
rel_abund_stool <- 100*stool_shared/unique(apply(stool_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_stool <- rel_abund_stool[,OTUs_stool_0_01]
OTU_list <- colnames(rel_abund_stool)

#Merge OTU data with mouse/exp data
#Predict_df <- merge(rf_meta_df, rel_abund_stool, by.x='RowNames', by.y='row.names')
#Predict_df <- Predict_df[,!names(Predict_df) %in% c('Mouse_ID','RowNames','cage_id')]

#predict day cfu by day 0 community
predict_df <- merge(meta_file[meta_file$day==2, c('Mouse_ID', 'log_cfu') ], stool_shared, by='Mouse_ID', by.y='row.names', all.y = T)
predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]

RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu),], importance = TRUE)

varImpPlot(RF)
ggplot(data = data.frame(actual=predict_df[,'log_cfu'],pred=predict(RF,predict_df)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
percent_var_expl <- RF$rsq[length(RF$rsq)]*100

#create function for random forest
run_rf_on <- function(DATAFRAME, RESPONSE, SAMPLE_SPLIT = 2/3,  OTUs = OTU_list){
set.seed(2)
     #
     Predictor_df <- as.data.frame(DATAFRAME[!DATAFRAME[,RESPONSE] %in% NA,c(RESPONSE,OTUs)])
     sample_set <- sample(seq_len(nrow(Predictor_df)), size = floor((SAMPLE_SPLIT)*nrow(Predictor_df)))
     test <- Predictor_df[-sample_set,]
     train <- Predictor_df[sample_set,] 
          RF <- randomForest(as.formula(paste(RESPONSE, '~ .')), data = train, importance = TRUE)
     R2 <- rSquared(test[,RESPONSE], test[,RESPONSE] - predict(RF,test))
     MSE <- mean((test[,RESPONSE] - predict(RF,test))^2)
     IMPORT <- importance(RF)
     FOREST <- RF
     PLOT<- ggplot(data = data.frame(actual=test[,RESPONSE],pred=predict(RF,test)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
     ##MEAN_DECREASE <- varImpPlot(RF)
     return(list(rf=FOREST,r2=R2,mse=MSE,import=IMPORT,plot=PLOT))
}
```

@@

##Can we predict moribundity with the community structure at Day 0?
```{r}
##Fix to be categorical, and not regression
Predict_early_euth_df <- Predict_df[!Predict_df$Euth_Early %in% NA,c('Euth_Early',OTU_list)]
rf_early_euth <- randomForest(as.factor(Euth_Early) ~ ., data = Predict_early_euth_df)
rf_early_euth
```

>>>>

# read in files
meta_file   <- read.table('data/process/human_CdGF_metadata.txt', sep = '\t',
                        header = T, row.names = 'group')
shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
                          sep='\t',header = T, row.names='Group')

# add log transformed CFU
meta_file$log_cfu <- log10(meta_file$cdiff_cfu + 1)
# remove cecum samples
meta_file <- meta_file[! meta_file$sample_type == 'cecum', ]

# subset shared
#Create df with relative abundances
stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
# use %in% instead of == because to ignore NA
# rename day 0 communities by mouse id
rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% 
                                               rownames(stool_shared) ]
rel_abund_stool <- 100 * stool_shared / unique(apply(stool_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
OTU_list <- colnames(rel_abund_stool)

#random forest regression model to determine Day 0 OTU correlated with increased CFU
seed <- 18
n_trees <- 2000
set.seed(seed)

#log_cfu_obs <- meta_file[meta_file$day %in% c(0:10), c('Mouse_ID', 'day', 'log_cfu') ]
#log_cfu_obs$predict <- NA
#log_cfu_obs[log_cfu_obs$day==2, ] <- predict(RF, rel_abund_stool)

importance_df <- data.frame(matrix(nrow = length(OTU_list), 
                                   ncol = max(meta_file$day, na.rm = T)),
                            row.names = OTU_list)
rsq_df <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
                         row.names = 'RSQ')
prediction_df <- data.frame(matrix(nrow = 0, ncol = 4))

for (i in 1:10){
  predict_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'log_cfu') ],
                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
  RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
                   importance = TRUE, ntree = n_trees)
  rsq_df[i] <- RF$rsq[length(RF$rsq)]
  # sort the features by their effect on % increase in the standard error
  importance_df[ , i] <- importance(RF)[,1]
  colnames(importance_df)[i] <- paste('day', i, 'importance', sep = '_')
# fit the model back on the datar and tack it to the end of the counts file
  random_forest_output <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day', 'log_cfu') ],
                              as.data.frame(predict(RF, rel_abund_stool)), 
                              by='Mouse_ID', by.y='row.names')
  colnames(random_forest_output)[3:4] <- c('observed', 'predicted')
  prediction_df <- rbind(prediction_df, random_forest_output)  
}

ggplot(data = prediction_df, aes(x=observed, y=predicted, color = day)) + geom_point()

rsq_df <- t(rsq_df)
rsq_df <- as.data.frame(cbind(rsq_df, day = c(1:10)))
ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_bar(stat = 'identity')

#create function for random forest
run_rf_on <- function(DATAFRAME, RESPONSE,  OTUs = OTU_list){
set.seed(2)
     #
     Predictor_df <- as.data.frame(DATAFRAME[!DATAFRAME[,RESPONSE] %in% NA,c(RESPONSE,OTUs)])
    RF <- randomForest(as.formula(paste(RESPONSE, '~ .')), data = train, importance = TRUE)
     R2 <- rSquared(test[,RESPONSE], test[,RESPONSE] - predict(RF,test))
     MSE <- mean((test[,RESPONSE] - predict(RF,test))^2)
     IMPORT <- importance(RF)
     FOREST <- RF
     PLOT<- ggplot(data = data.frame(actual=test[,RESPONSE],pred=predict(RF,test)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
     ##MEAN_DECREASE <- varImpPlot(RF)
     return(list(rf=FOREST,r2=R2,mse=MSE,import=IMPORT,plot=PLOT))
}

# let's plot the top 12 features
#n_features <- 12
#importance_subset <- importance_sorted[1:n_features]

# let's get Rsq for the subset
#rf_partial <- randomForest(log_cfu ~ ., data=predict_df[!is.na(predict_df$log_cfu),
                            c( 'log_cfu', names(importance_sorted)[1:n_features])], 
                           importance=TRUE, ntree=n_trees)
# write(paste("partial", n_features, rf_partial$rsq[n_trees], "\t"), file="data/process/random_forest.data", append=T)

@@

##Can we predict moribundity with the community structure at Day 0?
```{r}
##Fix to be categorical, and not regression
Predict_early_euth_df <- data.frame(Euth_Early = ifelse(meta_file[meta_file$day %in% 0,
                                                                  'Euthanized'] %in% 10,
                                                        'No','Yes'), 
                                    row.names = meta_file[meta_file$day %in% 0, 'Mouse_ID'])
Predict_early_euth_df <- merge(Predict_early_euth_df, rel_abund_stool, by='row.names')
Predict_early_euth_df <- Predict_early_euth_df[,!(colnames(Predict_early_euth_df) %in% "Row.names")]
rf_early_euth <- randomForest(as.factor(Euth_Early) ~ ., data = Predict_early_euth_df, 
                              importance = TRUE, ntree = n_trees)
# write(paste("binary", ncol(rel_abund_stool), rf_early_euth$err.rate[n_trees], "\t"), file="data/process/random_forest.data", append=T)
rf_early_euth
```
